Using numerical modelling to investigate temperature and wave exposure as drivers

### 2019-12-02 morphology_paper.Rmd
Final push for paper. output may be manually edited for some aspects but the changes will be incorportaed into the rmd

```{r include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(fig.align = "center")
```

# Introduction

@@

Given that kelps are cool-temperate organisms which are vulnerable to dislodgement under high wave exposure scenarios, increases in storm frequency and magnitude, and changes in ocean temperature, changes in these influences pose a direct threat to their survival; which is a concern considering that these variables are known to be affected by climate change[@Russo2014; @Reguero2019; @Weisse2010; @Rose2012; @Rummukainen2012]. Warmer temperatures reduces the resilience of kelp individuals to storm disturbance [@wernberg2010], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @wernberg2010]. A study by @Wernberg2003-mp investigated the morphology of *Ecklonia radiata* (C. Agardh) J. Agardh. in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australian coast. They found no correlation between spatial distance and morphological similarity; rather the morphology of the kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; @Mia2013; @Dubi1995].

@@

wave_boxplot_1 <- ggplot(data = wave_ann_meta, aes(x = region, y = hs_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y =0.6 ,label.x = 2.5, label = "p.signif") + 
  xlab("") + 
  ylab("") + 
  coord_flip() + 
  theme_bw() +
  scale_fill_grey(start = 0.4, end = 0.8) + 
  theme(axis.text.y=element_text(vjust=0.3)) 
ggsave("hs_plot.jpeg", plot = wave_boxplot_1, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

wave_boxplot_2 <- ggplot(data = wave_ann_meta, aes(x = region, y = tp_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y = 8.2, label.x = 2.5, label = "p.signif") + 
  xlab("") + 
  ylab("") +
  ylim(8,12) + 
  coord_flip() + 
  theme_bw() +
  scale_fill_grey(start = 0.4, end = 0.8) + 
  theme(axis.text.y=element_text(vjust=0.3))
ggsave("tp_plot.jpeg", plot = wave_boxplot_2, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

```{r wave rose, echo=FALSE, warning = FALSE, eval= TRUE, fig.align="center",fig.cap='\\label{fig:wave rose}The total and seasonal wave roses of the directional wave buoy just offshore from Cape Town is given form the years 2000 till end 2017 (the period for which directional wave spectra was available). Annual (top left), Austral summer (top right) and Austral winter (bottom) respectively'}

@@

img_wave_rose <- readPNG("figures/wave_rose.png")
grid.raster(img_wave_rose)

```{r wave power map, echo=FALSE, eval= TRUE, fig.cap='\\label{fig:wave power map}The total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length. Annual total (A), summer (B), winter (C), spring (D) and autumn (D).'}

wp_1 <- rasterGrob(as.raster(readPNG("figures/CapePen_Total.PNG")), interpolate = FALSE)
wp_2 <- rasterGrob(as.raster(readPNG("figures/CapePen_Summer.PNG")), interpolate = FALSE)
wp_3 <- rasterGrob(as.raster(readPNG("figures/CapePen_Winter.PNG")), interpolate = FALSE)
wp_4 <- rasterGrob(as.raster(readPNG("figures/CapePen_Spring.PNG")), interpolate = FALSE)
wp_5 <- rasterGrob(as.raster(readPNG("figures/CapePen_Autumn.PNG")), interpolate = FALSE)
wp_6 <- rasterGrob(as.raster(readPNG("figures/CapePen_Legend.PNG")), interpolate = FALSE)
grid.arrange(wp_1, wp_2, wp_3, wp_4, wp_5, wp_6, ncol = 3, nrow = 2)

#wave_power_map <- readPNG("figures/wave_power_all_labels.png")
#grid.raster(wave_power_map)

@@

ggsave("fig8A.jpeg", plot = LT1, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig8B.jpeg", plot = LT2, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig8C.jpeg", plot = LT3, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig8D.jpeg", plot = LT4, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig8E.jpeg", plot = LT5, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig8F.jpeg", plot = LT6, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig8G.jpeg", plot = LT7, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig8H.jpeg", plot = LT8, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig8I.jpeg", plot = LT9, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9A.jpeg", plot = LW1, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9B.jpeg", plot = LW2, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9C.jpeg", plot = LW3, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9D.jpeg", plot = LW4, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9E.jpeg", plot = LW5, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9F.jpeg", plot = LW6, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9G.jpeg", plot = LW7, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9H.jpeg", plot = LW8, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig9I.jpeg", plot = LW9, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10A.jpeg", plot = ET1, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10B.jpeg", plot = ET2, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10C.jpeg", plot = ET3, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10D.jpeg", plot = ET4, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10E.jpeg", plot = ET5, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10F.jpeg", plot = ET6, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10G.jpeg", plot = ET7, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10H.jpeg", plot = ET8, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10I.jpeg", plot = ET9, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig10J.jpeg", plot = ET10, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11A.jpeg", plot = EW1, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11B.jpeg", plot = EW2, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11C.jpeg", plot = EW3, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11D.jpeg", plot = EW4, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11E.jpeg", plot = EW5, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11F.jpeg", plot = EW6, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11G.jpeg", plot = EW7, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11H.jpeg", plot = EW8, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11I.jpeg", plot = EW9, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig11J.jpeg", plot = EW10, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12A.jpeg", plot = EST1, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12B.jpeg", plot = EST2, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12C.jpeg", plot = EST3, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12D.jpeg", plot = EST4, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12E.jpeg", plot = EST5, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12F.jpeg", plot = EST6, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12G.jpeg", plot = EST7, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12H.jpeg", plot = EST8, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12I.jpeg", plot = EST9, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig12J.jpeg", plot = EST10, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13A.jpeg", plot = ESW1, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13B.jpeg", plot = ESW2, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13C.jpeg", plot = ESW3, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13D.jpeg", plot = ESW4, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13E.jpeg", plot = ESW5, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13F.jpeg", plot = ESW6, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13G.jpeg", plot = ESW7, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13H.jpeg", plot = ESW8, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13I.jpeg", plot = ESW9, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

ggsave("fig13J.jpeg", plot = ESW10, dpi = 300, device = "jpeg", path = "extra/morphology_paper_prep/")

@@

@@

RC conceptualised the scope of the research reported in this paper, collected samples, undertook the majority of numerical and statistical analyses, made the first round of interpretation, and did the majority of the writing. AS and CR provided guidance in terms of  writing, statistical analyses, and the final draft of the paper, as well some writing in the relevant sections. RS aided in fieldwork and statistical analyses for the temperature data. TP performed the initial analyses and interpretation of the data.

@@

The research was funded by the South African National Research Foundation ([http://www.nrf.ac.za](http://www.nrf.ac.za)) programme "Thermal characteristics of the South African nearshore:implications for biodiversity" (CPRR14072378735).

>>>>

```{r include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(fig.align = "center")
```
# Introduction

@@

Given that kelps are cool-temperate organisms which are vulnerable to dislodgement under high wave exposure scenarios, increases in storm frequency and magnitude, and changes in ocean temperature, changes in these influences pose a direct threat to their survival; which is a concern considering that these variables are known to be affected by climate change[@Russo2014; @Reguero2019; @Weisse2010; @Rose2012; @Rummukainen2012]. Warmer temperatures reduces the resilience of kelp individuals to storm disturbance [@Wernberg2010b], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @Wernberg2010a]. A study by @Wernberg2003-mp investigated the morphology of *Ecklonia radiata* (C. Agardh) J. Agardh. in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australian coast. They found no correlation between spatial distance and morphological similarity; rather the morphology of the kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; @Mia2013; @Dubi1995].

@@

wave_boxplot_1 <- ggplot(data = wave_ann_meta, aes(x = region, y = hs_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y =0.6 ,label.x = 2.5, label = "p.signif") + 
  xlab("") + 
  ylab("") + 
  coord_flip() + 
  theme_bw() +
  scale_fill_grey(start = 0.4, end = 0.8) + 
  theme(axis.text.y=element_text(vjust=0.3))

wave_boxplot_2 <- ggplot(data = wave_ann_meta, aes(x = region, y = tp_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y = 8.2, label.x = 2.5, label = "p.signif") + 
  xlab("") + 
  ylab("") +
  ylim(8,12) + 
  coord_flip() + 
  theme_bw() +
  scale_fill_grey(start = 0.4, end = 0.8) + 
  theme(axis.text.y=element_text(vjust=0.3))

@@

```{r wave rose, echo=FALSE, warning = FALSE, eval= TRUE, fig.align="center",fig.cap='\\label{fig:wave rose}The total and seasonal wave roses of the directional wave buoy just offshore from Cape Town is given form the years 2000 till end 2017 (the period for which directional wave spectra was available). Total (A), Austral winter (B) and Austral summer (C) respectively'}

library(png)
library(jpeg)
library(grid)
library(gridExtra)

img_wave_rose_A <- rasterGrob(as.raster(readJPEG("figures/frontiers_fig_prep/fig_5A.jpg")))
img_wave_rose_B <- rasterGrob(as.raster(readJPEG("figures/frontiers_fig_prep/fig_5B.jpg")))
img_wave_rose_C <- rasterGrob(as.raster(readJPEG("figures/frontiers_fig_prep/fig_5C.jpg")))

ggarrange(img_wave_rose_A, img_wave_rose_B, img_wave_rose_C, ncol = 2, nrow = 2, labels = "AUTO")
```

```{r wave power map, echo=FALSE, eval= TRUE, fig.cap='\\label{fig:wave power map}The total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length. Total (A), summer (B), winter (C), spring (D) and autumn (E).'}

@@

wp_1 <- rasterGrob(as.raster(readPNG("figures/CapePen_Total.PNG")))
wp_2 <- rasterGrob(as.raster(readPNG("figures/CapePen_Summer.PNG")))
wp_3 <- rasterGrob(as.raster(readPNG("figures/CapePen_Winter.PNG")))
wp_4 <- rasterGrob(as.raster(readPNG("figures/CapePen_Spring.PNG")))
wp_5 <- rasterGrob(as.raster(readPNG("figures/CapePen_Autumn.PNG")))
wp_6 <- rasterGrob(as.raster(readPNG("figures/CapePen_Legend.PNG")))

ggarrange(wp_1, wp_2, wp_3, wp_4, wp_5, wp_6, ncol = 3, nrow = 2, labels = "AUTO", align = "hv")

#wave_power_map <- readPNG("figures/wave_power_all_labels.png")
#grid.raster(wave_power_map)
```

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

Kelps living in warmer environments have been shown to have a lower resilience to purbutation, such as storms [@Wernberg2010b]. The kelps in False Bay, may therefore be more vulnerable to dislodgment under high wave exposure scenarios compared to kelps on the western side of the Cape Peninsula, which experiences higher wave energy on a regular basis. In South Africa, the main contributing factors to the hydrodynamic environment are the Southern Annular Node (SAM) and El Nino Southern Oscillation (ENSO), while the latter is also the main source of variability [@Veitch2019]. These climate systems are expected to alter storm frequency and magnitude under future climate scenarios, which may increase the overall wave power around the Cape Peninsula. The magnitude-specific response of the kelp species to wave exposure in this study suggests there is a limit to morphological adaption in reducing the probability of dislodgement. Therefore, if this limit is reached and exceeded in future climate scenarios, kelp populations may risk being reduced over time. Furthermore, although recent studies suggest a cooling trend of coastal waters around South Africa [@Rouault2010-gi; @Smit2013a], the impact of marine heat waves may reduce the resilience of kelps to storms and high wave energy. Although this study identifies specific wave and temperature metrics that drive morphological adaptation and suggests a possible wave exposure threshold, it does not identify the metrics or magnitude of wave exposure required to dislodge kelp. Therefore, future studies should consider using advanced numerical methods to investigate the specific wave power needed to dislodge kelp, and ideally, given a specific set of morphological measures. These types of studies in conjunction with forecasting studies can be used to determine the vulnerability of kelp populations to future climate scenarios.

@@

RC conceptualised the scope of the research reported in this paper, collected samples, undertook the majority of numerical and statistical analyses, made the first round of interpretation, and did the majority of the writing. AS and CR provided guidance in terms of  writing, statistical analyses, and the final draft of the paper.

@@

The research was funded by the South African National Research Foundation ([http://www.nrf.ac.za](http://www.nrf.ac.za)). Funding number is CSRP170430229220
